---
title: "Practical 1: Amami rabbit Closed CMR"
subtitle: "Student handout summarising the data"
date: "Script last run on `r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: monochrome
    code_folding: hide
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: false
      smooth_scroll: false
    df_print: paged
---

Note that you are not allowed to use the figures in this document within your management plan.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
```

# Reminder of closed CMR models

The aim of closed CMR models is to estimate population size. To do so, we need to know how many individuals we failed to capture. Given we know how many animals we _did_ capture, we can add these two values together to get an estimate of total population size.

However, to estimate the number of individuals we failed to capture, we need to know what the capture probability was. If we know how likely we were to capture an animal, we know how _unlikely_ we were to _not_ capture an animal.

Closed CMR models estimate both the number of animals we did not capture (referred to as $f_0$, which stands for the frequency of animals, $f$, which were never capture $0$), in combination with the capture probability ($p$).

Given there may be a variety of processes that determined if an animal entered our trap or not (just like there may be multiple processes influencing your choice to drink coffee or not this morning), we have a suite of models that each represent different processes. These models include the fundamental forms:

* $M_o$, which posits that capture probability is best explained as a single, constant value (the Null model).

* $M_t$, which posits that the probability of capture varies over time.

* $M_b$, which posits that the probability of capturing an animal changes once it has been captured (e.g., trap happy or trap shy).

Note that these are only the fundamental models. They can be combined to propose, e.g., that there is a behaviour effect ($M_b$), that changes over time ($M_t$). 

See the [practical video](https://abdn.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=41c148c9-d2a1-4c90-8727-ac8e00ccf03a) for guidance on using `MARK` or for more information.

# Details of data collection

Beginning on the 3rd of July 1965, a capture-mark-recapture trapping session, lasting three days, occurred across the Amami island in a selection of sites known to have Amami rabbits present. Traps were baited with potato and carrots and were checked once every 24 hours. Researchers did not record any additional information.

The same data collection methodology was employed in 1983 in the same location.

In 2021, the methodology was altered to record the age of the rabbit at capture. These included one year olds, two year olds and adults (i.e., $\geq$ three years olds). Age was determined by trappers using the weight and sexual maturation of the captured animal.

# Practical objective

For each dataset, estimate the total population size. To do so will require that you construct a variety of models that posit different processes which influenced capture probability.

Begin with the 1965 dataset.

1. Follow the instructions in the practical video to get `MARK` prepared.

2. Construct your models that reflect different processes influencing capture probabilities.

3. Find your most parsimonious model using $\Delta$AIC.

4. For your most parsimonious model, save the output using notepad. Make notes now on any details of the model.

5. Save the `MARK` session

# Capture histories

## 1965

```{r}
# Set seed so that the random numbers are the same each time you run the script
set.seed(666)

# total number of rabbits
N <- 674

# number of days of trapping
t <- 3

# capture probabilities (must be same length as number of days)
p <- c(0.17, 0.34, 0.48)

# create empty matrix
cap <- matrix(nrow = N, ncol = t, data = NA)

# see how many rabbits are actually captured on each day
# Based on binomial distribution with probability p[day], do this for all rabbits (N) where there is only a single day to capture (1)
cap[,1] <- rbinom(N, 1, p[1])
cap[,2] <- rbinom(N, 1, p[2])
cap[,3] <- rbinom(N, 1, p[3])

# Combine and store it in a dataframe
cap1965 <- data.frame(
  CH = paste(cap[,1], cap[,2], cap[,3])
)

# Turn it into a frequency table
cap1965 <- cap1965 %>% 
  count(CH, sort = T)

# Remove rabbits never caught (this is what students are trying to estimate)
cap1965 <- cap1965[cap1965$CH != "0 0 0",]

# Adding comments for students

cap1965$comment <- NA

for(i in 1:nrow(cap1965)){
  if(cap1965$CH[i] == "1 1 1"){
    cap1965$comment[i] <- "/* Number of rabbits caught on day 1, 2, 3 */"
  } else if(cap1965$CH[i] == "1 1 0") {
    cap1965$comment[i] <- "/* Number of rabbits caught on day 1, 2 */"
  } else if(cap1965$CH[i] == "0 1 1") {
    cap1965$comment[i] <- "/* Number of rabbits caught on day 2, 3 */"
  } else if(cap1965$CH[i] == "1 0 1") {
    cap1965$comment[i] <- "/* Number of rabbits caught on day 1, 3 */"
  } else if(cap1965$CH[i] == "1 0 0") {
    cap1965$comment[i] <- "/* Number of rabbits caught on day 1 */"
  } else if(cap1965$CH[i] == "0 1 0") {
    cap1965$comment[i] <- "/* Number of rabbits caught on day 2 */"
  } else if(cap1965$CH[i] == "0 0 1") {
    cap1965$comment[i] <- "/* Number of rabbits caught on day 3 */"
  }
}

ggplot(cap1965) +
  geom_col(aes(x = reorder(CH, -n), y = n)) +
  labs(x = "Capture history (day1 | day2 | day3)",
       y = "Number of unique rabbits caught",
       title = "Amami rabbit capture histories in 1965",
       caption = paste("Total number of rabbits caught =", sum(as.numeric(cap1965$n)), "\n",
                       "Number of rabbits never caught (f0) = ?", "\n",
                       "Total number of rabbits =", sum(as.numeric(cap1965$n)), "+ f0 = ?")) +
  theme_bw() +
  theme_minimal()
```

## 1983

```{r}
# Set seed so that the random numbers are the same each time you run the script
set.seed(666)

# total number of rabbits
N <- 421

# number of days of trapping
t <- 3

# capture probabilities (must be same length as number of days)
p <- c(0.07, 0.24, 0.38)

# recapture probability (must be of length t - 1)
pr <- c(0.64, 0.87)

# create empty matrix
cap <- matrix(nrow = N, ncol = t, data = NA)

# see how many rabbits are actually captured on each day
# Based on binomial distribution with probability p[day], do this for all rabbits (N) where there is only a single day to capture (1)
cap[,1] <- rbinom(N, 1, p[1])

# For day 2, use a for loop to go through each individual (i)
# If rabbit [i] had not been caught on day 1 (cap[i,1] == 0)
  # Then it uses the capture prob for day 2 p[2]
  # But if it was caught, then use recapture prob for day 2 (pr[1])
for(i in 1:nrow(cap)){
  cap[i,2] <- ifelse(cap[i,1] == 0, rbinom(N, 1, p[2]), rbinom(N, 1, pr[1]))
}
# Repeat above loop but for last day
for(i in 1:nrow(cap)){
  cap[i,3] <- ifelse(cap[i,1] == 0 & cap[i,2] == 0, rbinom(N, 1, p[3]), rbinom(N, 1, pr[2]))
}

# Combine and store it in a dataframe
cap1983 <- data.frame(
  CH = paste(cap[,1], cap[,2], cap[,3])
)

# Turn it into a frequency table
cap1983 <- cap1983 %>% 
  count(CH, sort = T)

# Remove rabbits never caught (this is what students are trying to estimate)
cap1983 <- cap1983[cap1983$CH != "0 0 0",]

# Adding comments for students

cap1983$comment <- NA

for(i in 1:nrow(cap1983)){
  if(cap1983$CH[i] == "1 1 1"){
    cap1983$comment[i] <- "/* Number of rabbits caught on day 1, 2, 3 */"
  } else if(cap1983$CH[i] == "1 1 0") {
    cap1983$comment[i] <- "/* Number of rabbits caught on day 1, 2 */"
  } else if(cap1983$CH[i] == "0 1 1") {
    cap1983$comment[i] <- "/* Number of rabbits caught on day 2, 3 */"
  } else if(cap1983$CH[i] == "1 0 1") {
    cap1983$comment[i] <- "/* Number of rabbits caught on day 1, 3 */"
  } else if(cap1983$CH[i] == "1 0 0") {
    cap1983$comment[i] <- "/* Number of rabbits caught on day 1 */"
  } else if(cap1983$CH[i] == "0 1 0") {
    cap1983$comment[i] <- "/* Number of rabbits caught on day 2 */"
  } else if(cap1983$CH[i] == "0 0 1") {
    cap1983$comment[i] <- "/* Number of rabbits caught on day 3 */"
  }
}

ggplot(cap1983) +
  geom_col(aes(x = reorder(CH, -n), y = n)) +
  labs(x = "Capture history (day1 | day2 | day3)",
       y = "Number of unique rabbits caught",
       title = "Amami rabbit capture histories in 1983",
       caption = paste("Total number of rabbits caught =", sum(as.numeric(cap1983$n)), "\n",
                       "Number of rabbits never caught (f0) = ?", "\n",
                       "Total number of rabbits =", sum(as.numeric(cap1983$n)), "+ f0 = ?")) +
  theme_bw() +
  theme_minimal()
```

## 2021

### One year olds

```{r}
# Set seed so that the random numbers are the same each time you run the script
set.seed(666)

# total number of rabbits
N <- 83

# number of days of trapping
t <- 3

# capture probabilities (must be same length as number of days)
p <- c(0.36, 0.36, 0.36)

# recapture probability (must be of length t - 1)
pr <- c(0.42, 0.54)

# create empty matrix
cap <- matrix(nrow = N, ncol = t, data = NA)

# see how many rabbits are actually captured on each day
# Based on binomial distribution with probability p[day], do this for all rabbits (N) where there is only a single day to capture (1)
cap[,1] <- rbinom(N, 1, p[1])
# For day 2, use a for loop to go through each individual (i)
# If rabbit [i] had not been caught on day 1 (cap[i,1] == 0)
  # Then it uses the capture prob for day 2 p[2]
  # But if it was caught, then use recapture prob for day 2 (pr[1])
for(i in 1:nrow(cap)){
  cap[i,2] <- ifelse(cap[i,1] == 0, rbinom(N, 1, p[2]), rbinom(N, 1, pr[1]))
}
for(i in 1:nrow(cap)){
  cap[i,3] <- ifelse(cap[i,1] == 0 & cap[i,2] == 0, rbinom(N, 1, p[3]), rbinom(N, 1, pr[2]))
}

# Combine and store it in a dataframe
cap2022_juv <- data.frame(
  CH = paste(cap[,1], cap[,2], cap[,3])
)

# Turn it into a frequency table
cap2022_juv <- cap2022_juv %>% 
  count(CH, sort = T)

# Remove rabbits never caught (this is what students are trying to estimate)
cap2022_juv <- cap2022_juv[cap2022_juv$CH != "0 0 0",]

# Adding comments for students

cap2022_juv$comment <- NA

for(i in 1:nrow(cap2022_juv)){
  if(cap2022_juv$CH[i] == "1 1 1"){
    cap2022_juv$comment[i] <- "/* Number of rabbits caught on day 1, 2, 3 */"
  } else if(cap2022_juv$CH[i] == "1 1 0") {
    cap2022_juv$comment[i] <- "/* Number of rabbits caught on day 1, 2 */"
  } else if(cap2022_juv$CH[i] == "0 1 1") {
    cap2022_juv$comment[i] <- "/* Number of rabbits caught on day 2, 3 */"
  } else if(cap2022_juv$CH[i] == "1 0 1") {
    cap2022_juv$comment[i] <- "/* Number of rabbits caught on day 1, 3 */"
  } else if(cap2022_juv$CH[i] == "1 0 0") {
    cap2022_juv$comment[i] <- "/* Number of rabbits caught on day 1 */"
  } else if(cap2022_juv$CH[i] == "0 1 0") {
    cap2022_juv$comment[i] <- "/* Number of rabbits caught on day 2 */"
  } else if(cap2022_juv$CH[i] == "0 0 1") {
    cap2022_juv$comment[i] <- "/* Number of rabbits caught on day 3 */"
  }
}

ggplot(cap2022_juv) +
  geom_col(aes(x = reorder(CH, -n), y = n)) +
  labs(x = "Capture history (day1 | day2 | day3)",
       y = "Number of unique juvenile rabbits caught",
       title = "Juvenile Amami rabbit capture histories in 2022",
       caption = paste("Total number of one year old rabbits caught =", sum(as.numeric(cap2022_juv$n)), "\n",
                       "Number of one year old rabbits never caught (f0) = ?", "\n",
                       "Total number of one year old rabbits =", sum(as.numeric(cap2022_juv$n)), "+ f0 = ?")) +
  theme_bw() +
  theme_minimal()
```

### Two year olds

```{r}
# Set seed so that the random numbers are the same each time you run the script
set.seed(666)

# total number of rabbits
N <- 59

# number of t of trapping
t <- 3

# capture probabilities (must be same length as number of t)
p <- c(0.46, 0.23, 0.13)

# recapture probability (must be of length t - 1)
pr <- c(0.39, 0.48)

# create empty matrix
cap <- matrix(nrow = N, ncol = t, data = NA)

# see how many rabbits are actually captured on each day
# Based on binomial distribution with probability p[day], do this for all rabbits (N) where there is only a single day to capture (1)
cap[,1] <- rbinom(N, 1, p[1])
# For day 2, use a for loop to go through each individual (i)
# If rabbit [i] had not been caught on day 1 (cap[i,1] == 0)
  # Then it uses the capture prob for day 2 p[2]
  # But if it was caught, then use recapture prob for day 2 (pr[1])
for(i in 1:nrow(cap)){
  cap[i,2] <- ifelse(cap[i,1] == 0, rbinom(N, 1, p[2]), rbinom(N, 1, pr[1]))
}
for(i in 1:nrow(cap)){
  cap[i,3] <- ifelse(cap[i,1] == 0 & cap[i,2] == 0, rbinom(N, 1, p[3]), rbinom(N, 1, pr[2]))
}

# Combine and store it in a dataframe
cap2022_suba <- data.frame(
  CH = paste(cap[,1], cap[,2], cap[,3])
)

# Turn it into a frequency table
cap2022_suba <- cap2022_suba %>% 
  count(CH, sort = T)

# Remove rabbits never caught (this is what students are trying to estimate)
cap2022_suba <- cap2022_suba[cap2022_suba$CH != "0 0 0",]

# Adding comments for students

cap2022_suba$comment <- NA

for(i in 1:nrow(cap2022_juv)){
  if(cap2022_suba$CH[i] == "1 1 1"){
    cap2022_suba$comment[i] <- "/* Number of rabbits caught on day 1, 2, 3 */"
  } else if(cap2022_suba$CH[i] == "1 1 0") {
    cap2022_suba$comment[i] <- "/* Number of rabbits caught on day 1, 2 */"
  } else if(cap2022_suba$CH[i] == "0 1 1") {
    cap2022_suba$comment[i] <- "/* Number of rabbits caught on day 2, 3 */"
  } else if(cap2022_suba$CH[i] == "1 0 1") {
    cap2022_suba$comment[i] <- "/* Number of rabbits caught on day 1, 3 */"
  } else if(cap2022_suba$CH[i] == "1 0 0") {
    cap2022_suba$comment[i] <- "/* Number of rabbits caught on day 1 */"
  } else if(cap2022_suba$CH[i] == "0 1 0") {
    cap2022_suba$comment[i] <- "/* Number of rabbits caught on day 2 */"
  } else if(cap2022_suba$CH[i] == "0 0 1") {
    cap2022_suba$comment[i] <- "/* Number of rabbits caught on day 3 */"
  }
}

ggplot(cap2022_suba) +
  geom_col(aes(x = reorder(CH, -n), y = n)) +
  labs(x = "Capture history (day1 | day2 | day3)",
       y = "Number of unique subadult rabbits caught",
       title = "Subadult Amami rabbit capture histories in 2022",
       caption = paste("Total number of two year old rabbits caught =", sum(as.numeric(cap2022_suba$n)), "\n",
                       "Number of two year old rabbits never caught (f0) = ?", "\n",
                       "Total number of two year old rabbits =", sum(as.numeric(cap2022_suba$n)), "+ f0 = ?")) +
  theme_bw() +
  theme_minimal()
```

### Adults

```{r}
# Set seed so that the random numbers are the same each time you run the script
set.seed(666)

# total number of rabbits
N <- 106

# number of days of trapping
t <- 3

# capture probabilities (must be same length as number of days)
p <- c(0.26, 0.38, 0.43)

# recapture probability (must be of length t - 1)
pr <- c(0.56, 0.56)

# create empty matrix
cap <- matrix(nrow = N, ncol = t, data = NA)

# see how many rabbits are actually captured on each day
# Based on binomial distribution with probability p[day], do this for all rabbits (N) where there is only a single day to capture (1)
cap[,1] <- rbinom(N, 1, p[1])
# For day 2, use a for loop to go through each individual (i)
# If rabbit [i] had not been caught on day 1 (cap[i,1] == 0)
  # Then it uses the capture prob for day 2 p[2]
  # But if it was caught, then use recapture prob for day 2 (pr[1])
for(i in 1:nrow(cap)){
  cap[i,2] <- ifelse(cap[i,1] == 0, rbinom(N, 1, p[2]), rbinom(N, 1, pr[1]))
}
for(i in 1:nrow(cap)){
  cap[i,3] <- ifelse(cap[i,1] == 0 & cap[i,2] == 0, rbinom(N, 1, p[3]), rbinom(N, 1, pr[2]))
}

# Combine and store it in a dataframe
cap2022_adult <- data.frame(
  CH = paste(cap[,1], cap[,2], cap[,3])
)

# Turn it into a frequency table
cap2022_adult <- cap2022_adult %>% 
  count(CH, sort = T)

# Remove rabbits never caught (this is what students are trying to estimate)
cap2022_adult <- cap2022_adult[cap2022_adult$CH != "0 0 0",]

# Adding comments for students

cap2022_adult$comment <- NA

for(i in 1:nrow(cap2022_juv)){
  if(cap2022_adult$CH[i] == "1 1 1"){
    cap2022_adult$comment[i] <- "/* Number of rabbits caught on day 1, 2, 3 */"
  } else if(cap2022_adult$CH[i] == "1 1 0") {
    cap2022_adult$comment[i] <- "/* Number of rabbits caught on day 1, 2 */"
  } else if(cap2022_adult$CH[i] == "0 1 1") {
    cap2022_adult$comment[i] <- "/* Number of rabbits caught on day 2, 3 */"
  } else if(cap2022_adult$CH[i] == "1 0 1") {
    cap2022_adult$comment[i] <- "/* Number of rabbits caught on day 1, 3 */"
  } else if(cap2022_adult$CH[i] == "1 0 0") {
    cap2022_adult$comment[i] <- "/* Number of rabbits caught on day 1 */"
  } else if(cap2022_adult$CH[i] == "0 1 0") {
    cap2022_adult$comment[i] <- "/* Number of rabbits caught on day 2 */"
  } else if(cap2022_adult$CH[i] == "0 0 1") {
    cap2022_adult$comment[i] <- "/* Number of rabbits caught on day 3 */"
  }
}

ggplot(cap2022_adult) +
  geom_col(aes(x = reorder(CH, -n), y = n)) +
  labs(x = "Capture history (day1 | day2 | day3)",
       y = "Number of unique adult rabbits caught",
       title = "Adult Amami rabbit capture histories in 2022",
       caption = paste("Total number of two year old rabbits caught =", sum(as.numeric(cap2022_adult$n)), "\n",
                       "Number of two year old rabbits never caught (f0) = ?", "\n",
                       "Total number of two year old rabbits =", sum(as.numeric(cap2022_adult$n)), "+ f0 = ?")) +
  theme_bw() +
  theme_minimal()
```